The Precambrian paleogeography of Laurentia

### 2021-03-06 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add Bruce Buffett acknowledgement
small manuscript edits

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 billion years ago; Ga), as well as regions of Paleoproterozoic and Mesoproterozoic crustal growth that post-date this assembly (Fig. \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}). That the vast majority of the modern-day continent of North America is a single Precambrian craton without major differential motion between constituent provinces and relatively minor crustal growth over the past billion years is exceptional in comparison to Earth's other continents. In contrast, South America and Africa are the product of the amalgamation of multiple Proterozoic cratons that obtained their relative positions during the formation of Gondwana ca. 0.6 Ga \citep{Goscombe2019a}. Eurasia's constituent Proterozoic cratons have an even more recent history of amalgmation with the North China and South China cratons not arriving in their present relative position until ca. 0.15 Ga \citep{Van-der-Voo2015a,Torsvik2017a}. The longevity of a large intact Laurentia makes its position a critical part of global paleogeographic models since its assembly. Rich geologic, paleomagnetic and geochronologic data give deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.

@@

\subsubsection{Laurentia's initial formation}
Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig. \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}). Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of the major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

Crustal growth also progressed in the Paleoproterozoic through accretionary orogenesis. This accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was the peripheral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca. 1.86 to 1.82 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Schulz2007a}). Firm evidence of the end of the Penokean orogeny comes from the ca. 1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.

@@

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth, as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny (Fig. \ref{fig:tectonic_history}). Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Potentially associated with the Yavapai orogeny is the accretion of the Mojave province of southwestern Laurentia (Fig. \ref{fig:Laurentia_map}). The Mojave province comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}. It is unclear whether the Mojave province should be considered to be a distinct province of Archean lithosphere akin to other Archean provinces or rather it should dominantly be considered to be Yavapai arc terrane built upon minor fragments of Archean lithosphere \citep{Whitmeyer2007a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeastern Canada in Labrador where it is called the Labradorian orogeny and is interpreted to have been active from ca. 1.71 to 1.60 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).

@@

Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration. This configuration had originally been proposed by \citet{Patchett1978a} and \citet{Piper1980a} largely on the basis of paleomagnetic pole comparisons. This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}. In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca. 1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia \citep{Karlstrom2001a}. Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}. Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca. 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca. 1.12 Ga if a virtual geomagnetic pole from the Salla dike of northeastern Finland is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}. These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.

@@

In terms of the orogenic timing, the main phase of the Sveconorwegian orogeny from ca. 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}. It is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}. \cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca. 1.0 Ga. However, in a recent critical review of the constraints, the high pressure nature of metamorphism and the cratonward propagation of orogenesis (in contrast to older accretionary orogenesis) led \cite{Stephens2020a} to favor the interpretation that the Sveconorwegian orogen is a record of prolonged continent-continent collision initiating ca. 1.06 Ga. In this model, Baltica could have been on the same plate as Laurentia during the rapid late Mesoporoterozoic motion leading up to collisional Grenvillian orogenesis and the associated assembly of Rodinia (Fig. \ref{fig:Grenville_reconstructions}).

Intrusive granitoids and extrusive calc-alkaline volcanics with early Tonian Period ages between 975 to 915 Ma are present in East Greenland and related Arctic terranes including northeast Svalbard \citep{McClelland2019a}. These intrusive and volcanic rocks are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a} that has been hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}. The need for the East Greenland margin to have been active in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation that would have severed the NENA connection \citep{Cawood2010a}. Such a clockwise rotation of Baltica relative to Laurentia from a Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador (northeast Canada) is implemented in many paleogeographic models (e.g. \citealp{Evans2009a} and as shown in Fig. \ref{fig:Grenville_reconstructions}). This position relative to Laurentia results in a joint Laurentia-Baltica APWP with large oscillations implied by Baltica's APWP. These oscillations would imply that following the Grenville loop, Laurentia moved northward across the equator and then back south to a similar Grenville loop position prior to the ca. 780 Ma Laurentia poles \citep{Evans2015a,Fairchild2017a}.  A challenge with this model in terms of regional tectonics, is that it is unclear what conjugate lithosphere rifted from East Greenland leading to the thick sedimentary succession interpreted to have been deposited on a thermally-subsiding passive margin from ca. 850 Ma into the Cryogenian (Fig. \ref{fig:tectonic_history}; \citealp{Maloof2006a}). \cite{Malone2014a} interpreted this Neoproterozoic basin development along the East Greenland margin to have been the result of extension in a back-arc setting.

@@

High-quality paleomagnetic poles constrain the coeval ca. 1109 Ma Umkondo large igneous province of the Kalahari craton and the early flood basalts of the Midcontinent Rift of Laurentia to have been separated by more than 50\textdegree\ of latitude at the time they were emplaced. These poles reconstruct the craton margins as having been separated by more than 30\textdegree\ of latitude \citep{Swanson-Hysell2015a}. These data make it difficult to envision a shared origin of magmatism and pose a challenge to approaches that seek to reconstruct paleogeography on the basis on ages of LIPs alone. However, with the subsequent rapid motion of Laurentia to low latitudes (Figs. \ref{fig:Laurentia_paleolatitude} and \ref{fig:Grenville_reconstructions}), it is possible that the Kalahari was a conjugate craton to the (south)eastern margin of Laurentia during the time of the Grenvillian orogeny. This conjugate relationship based on the interpretation that the Namaqua-Natal belt in southern Kalahari records late Mesoproterozoic collisional orogenesis was  proposed in \cite{Hoffman1991a} and is implemented in many reconstructions of Rodinia (e.g. \citealp{Li2008a}). Whether the Grenvillian margin of Laurentia and the Namaqua-Natal belt of Kalahari faced one another and could have been conjugates can be evaluated by paired paleomagnetic and geochronologic data sets from the Umkondo LIP and the Midcontinent Rift. The preferred interpretation of \cite{Swanson-Hysell2015a} and \cite{Kasbohm2015a} is that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g. \citealp{Swanson-Hysell2014a}) such that Namaqua-Natal margin faced the Grenvillian margin of Laurentia (Fig. \ref{fig:Grenville_reconstructions}). The late Mesoproterozoic apparent polar wander paths for Laurentia and Kalahari are consistent with them becoming conjoined as in Figure \ref{fig:Grenville_reconstructions} \citep{Swanson-Hysell2015a}. The record of the Namaqua belt wherein there is granitoid plutonism and arc accretion up to ca. 1090 followed by peak granulite metamorphism ca. 1065 to 1045 Ma \citep{Diener2013a, Spencer2015a} is consistent with a scenario wherein Kalahari was on the upper plate that collided with Laurentia at the time of the Ottawan phase of Grenvillian orogenesis following subduction of oceanic lithosphere associated with an intervening ocean. If they indeed became conjoined in Rodinia as in Figure \ref{fig:Grenville_reconstructions}, the separation of Kalahari from Laurentia may have initiated ca. 795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and subsequent Tonian volcanics in southeastern Laurentia. This late Tonian timing for separation could be  consistent with a position along the southeastern Laurentia margin given the ca. 750 Ma dates for the Mount Rogers volcanics and associated rift-related sediments \citep{Aleinikoff1995a,MacLennan2020a} --- although subsequent Ediacaran rifting and Cambrian thermal subsidence in the region would need to be attributed to the rifting of microcontinents such as Arequipa and Cuyania \citep{Escayola2011a, Martin2019a}.

@@

The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Zhang2021a}. While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} proposed that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwestern margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. A challenge with interpreting these North China poles as primary is that they include data from limestones that would reconstruct them to have been deposited at very high latitude ($>$80\textdegree\ for the limestones of the lower member of the Nanfen Formation). Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes due to the temperature dependence of carbonate saturation. As additional support for a North China--Laurentia connection, \cite{Zhao2019a} point to similarities in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins as supporting this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the 1237 $\pm$ 5 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma Narssaq Gabbro and Hvidaal dike poles during which it ascends to the Logan Loop (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). In its reconstructed position, the ca. 1220 Ma North China pole is close to this gap such that it is possible that the APWPs are consistent and that North China arrived on the northwestern Laurentia margin prior to 1220 Ma. In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwestern Laurentia.

@@

\noindent\textbf{Granite-rhyolite province } A geologic province that comprises widespread Mesoproterozoic rhyolite and granite extending from southeastern Ontario, Canada to west Texas, USA.

\noindent\textbf{Grenvillian orogen } An orogen resulting from collision orogeny between Laurentia and conjugate continent(s) near the end of the Mesoproterozoic.

@@

\noindent\textbf{juvenile } An adjuective referring to rocks that have formed through melt recently extracted from the mantle.

@@

\noindent\textbf{Nuna } A hypothesized supercontinent interpreted to have formed late in the Paleoproterozoic era and to have broken apart in the middle of the Mesoproterozoic.

@@

\noindent\textbf{Rodinia } A hypothesized supercontinent interpreted to have formed late in the Mesoproterozoic era at the time of the Grenvillian orogeny and to have broken apart in the middle of the Neoproterozoic.

>>>>

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 billion years ago; Ga), as well as regions of Paleoproterozoic and Mesoproterozoic crustal growth that post-date this assembly (Fig. \ref{fig:Laurentia_map}; \citealp{Hoffman1989c, Whitmeyer2007a}). That the vast majority of the modern-day continent of North America is a single Precambrian craton without major differential motion between constituent provinces and relatively minor crustal growth over the past billion years is exceptional in comparison to Earth's other continents. In contrast, South America and Africa are products of the amalgamation of multiple Proterozoic cratons that obtained their relative positions during the formation of Gondwana ca. 0.6 Ga \citep{Goscombe2019a}. Eurasia's constituent Proterozoic cratons have an even more recent history of amalgmation with the North China and South China cratons not arriving in their present relative position until ca. 0.15 Ga \citep{Van-der-Voo2015a,Torsvik2017a}. The longevity of a large intact Laurentia makes its position a critical part of global paleogeographic models since its assembly. Rich geologic, paleomagnetic and geochronologic data provide deep insight into Laurentia's tectonic history and paleogeographic journey that is the focus of this chapter.

@@

\subsubsection{Laurentia's initial formation}
Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig. \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}). Terminal collision recorded in the Trans-Hudson orogen is estimated to have occured ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of the major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

Growth of Laurentia in the Paleoproterozoic also occured through accretionary orogenesis. This accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was the peripheral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca. 1.86 to 1.82 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Schulz2007a}). Firm evidence of the end of the Penokean orogeny comes from the ca. 1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.

@@

Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a shift in the predominant style of Laurentia's growth, as subsequent crustal growth occurred dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean provinces (\citealp{Whitmeyer2007a}; Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). Determining the extent of these belts is complicated by poor exposure of them in the midcontinent relative to the exposure of the Archean provinces throughout the Canadian shield. Major growth of Laurentia following the amalgamation of these Archean provinces occurred associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny (Fig. \ref{fig:tectonic_history}). Yavapai orogenesis is interpreted to have resulted from the accretion of a series of arc terranes that collided with each other and Laurentia \citep{Karlstrom2001a}. Potentially associated with the Yavapai orogeny is the accretion of the Mojave province of southwestern Laurentia (Fig. \ref{fig:Laurentia_map}). The Mojave province comprises Paleoproterozoic gneiss that is interpreted based on isotopic data to include reworked Archean lithologies \citep{Bennett1987a}. However, it remains unclear whether the Mojave province should be considered a Archean province or a Yavapai arc terrane built upon minor fragments of Archean lithosphere \citep{Whitmeyer2007a}. Yavapai accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stabilized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65 to 1.60 Ga Mazatzal orogeny and associated plutonism led to further crustal growth in the latest Paleoproterozoic \citep{Karlstrom1988a}. Yavapai to Mazatzal-age accretionary orogenesis extended from the southwestern USA to northeastern Canada in Labrador where it is called the Labradorian orogeny and is interpreted to have been active from ca. 1.71 to 1.60 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Gower1992a, Gower2008b}).

@@

Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration. This configuration had originally been proposed by \citet{Patchett1978a} and \citet{Piper1980a} largely on the basis of paleomagnetic pole comparisons. This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}. In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca. 1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia \citep{Karlstrom2001a}. Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}. Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca. 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca. 1.12 Ga if a virtual geomagnetic pole (that is a pole position calculated from paleomagnetic data of a single cooling unit) from the Salla dike of northeastern Finland is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}. These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.

@@

In terms of the orogenic timing, the main phase of the Sveconorwegian orogeny from ca. 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}. There is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}. \cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca. 1.0 Ga. However, in a recent critical review of the constraints, the high pressure nature of metamorphism and the cratonward propagation of orogenesis (in contrast to older accretionary orogenesis) led \cite{Stephens2020a} to favor the interpretation that the Sveconorwegian orogen is a record of prolonged continent-continent collision initiating ca. 1.06 Ga. In this model, Baltica could have been on the same plate as Laurentia during the rapid late Mesoporoterozoic motion leading up to collisional Grenvillian orogenesis and the associated assembly of Rodinia (Fig. \ref{fig:Grenville_reconstructions}).

Early Tonian intrusive granitoids and extrusive calc-alkaline volcanics in East Greenland and related Arctic terranes including northeast Svalbard, with ages between 975 to 915 Ma \citep{McClelland2019a}, are interpreted to have formed within a magmatic arc or in a syn-collisional setting \citep{Johansson1999a}. This magmatic activity is hypothesized to be the result of subduction and accretionary orogenesis termed the Valhalla orogeny \citep{Cawood2010a}. This geological evidence for an active margin in East Greenland in the earliest Neoproterozoic has motivated a tectonic model where Baltica rifted off East Greenland in the late Mesoproterozoic while staying proximal to Laurentia via a clockwise rotation that would have severed the NENA connection \citep{Cawood2010a}. Such a clockwise rotation of Baltica relative to Laurentia from the earlier Paleoproterozoic-Mesoproterozoic NENA configuration with a shared Labradorian to Gothian orogenic belt to one where the Sveconorwegian orogen is close to the Grenvillian orogen in Labrador (northeast Canada) is implemented in many paleogeographic models (e.g. \citealp{Evans2009a} and as shown in Fig. \ref{fig:Grenville_reconstructions}). This position relative to Laurentia results in a joint Laurentia-Baltica APWP with large oscillations implied by Baltica poles. These oscillations would imply that following the Grenville loop, Laurentia moved northward across the equator and then back south to a similar Grenville loop position prior to the ca. 780 Ma Laurentia poles \citep{Evans2015a,Fairchild2017a}.  A challenge with this model in terms of regional tectonics, is that it is unclear what conjugate lithosphere rifted from East Greenland leading to the thick sedimentary succession interpreted to have been deposited on a thermally-subsiding passive margin from ca. 850 Ma into the Cryogenian (Fig. \ref{fig:tectonic_history}; \citealp{Maloof2006a}). \cite{Malone2014a} interpreted this Neoproterozoic basin development along the East Greenland margin to have been the result of extension in a back-arc setting.

@@

High-quality paleomagnetic poles constrain the coeval ca. 1109 Ma Umkondo large igneous province of the Kalahari craton and the early flood basalts of the Midcontinent Rift of Laurentia to have been separated by more than 50\textdegree\ of latitude at the time they were emplaced. These poles reconstruct the craton margins as having been separated by more than 30\textdegree\ of latitude \citep{Swanson-Hysell2015a}. These data make it difficult to envision a shared origin of magmatism and pose a challenge to approaches that seek to reconstruct paleogeography on the basis on ages of LIPs alone. However, with the subsequent rapid motion of Laurentia to low latitudes (Figs. \ref{fig:Laurentia_paleolatitude} and \ref{fig:Grenville_reconstructions}), it is possible that the Kalahari was a conjugate craton to the (south)eastern margin of Laurentia during the time of the Grenvillian orogeny. This conjugate relationship based on the interpretation that the Namaqua-Natal belt in southern Kalahari records late Mesoproterozoic collisional orogenesis was  proposed in \cite{Hoffman1991a} and is implemented in many reconstructions of Rodinia (e.g. \citealp{Li2008a}). Whether the Grenvillian margin of Laurentia and the Namaqua-Natal belt of Kalahari faced one another and could have been conjugates can be evaluated by paired paleomagnetic and geochronologic data sets from the Umkondo LIP and the Midcontinent Rift. The preferred interpretation of \cite{Swanson-Hysell2015a} and \cite{Kasbohm2015a} is that sites with northerly declinations from the Umkondo Province correspond to the reversed polarity directions from the early magmatic stage in the Midcontinent Rift (e.g. \citealp{Swanson-Hysell2014a}) such that Namaqua-Natal margin faced the Grenvillian margin of Laurentia (Fig. \ref{fig:Grenville_reconstructions}). The late Mesoproterozoic apparent polar wander paths for Laurentia and Kalahari are consistent with them becoming conjoined as in Figure \ref{fig:Grenville_reconstructions} \citep{Swanson-Hysell2015a}. The record of the Namaqua belt wherein there is granitoid plutonism and arc accretion up to ca. 1090 followed by peak granulite metamorphism ca. 1065 to 1045 Ma \citep{Diener2013a, Spencer2015a} is consistent with a scenario wherein Kalahari was on the upper plate that collided with Laurentia at the time of the Ottawan phase of Grenvillian orogenesis following subduction of oceanic lithosphere associated with an intervening ocean. If they indeed became conjoined in Rodinia as in Figure \ref{fig:Grenville_reconstructions}, the separation of Kalahari from Laurentia may have initiated ca. 795 Ma heralded by the emplacement of the Gannakouriep diabase dike swarm \citep{Rioux2010a, DeKock2021a} and volcanic rocks in southeastern Laurentia (e.g., the ca. 750 Ma Mount Rogers volcanics and associated rift-related sediments \cite{Aleinikoff1995a, MacLennan2020a}). This late Tonian timing for separation could be consistent with a position along the southeastern Laurentia margin, although subsequent Ediacaran rifting and Cambrian thermal subsidence in the region would need to be attributed to the rifting of microcontinents such as Arequipa and Cuyania \citep{Escayola2011a, Martin2019a}.

@@

The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Zhang2021a}. While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} proposed that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwestern margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. A challenge with interpreting these North China poles as primary is that they include data from limestones that would reconstruct them to have been deposited at very high latitude ($>$80\textdegree\ for the limestones of the lower member of the Nanfen Formation). Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes due to the temperature dependence of carbonate saturation. As additional support for a North China--Laurentia connection, \cite{Zhao2019a} pointed to similarities in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins as supporting this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the 1237 $\pm$ 5 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma Narssaq Gabbro and Hvidaal dike poles during which it ascends to the Logan Loop (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). In its reconstructed position, the ca. 1220 Ma North China pole is close to this gap such that it is possible that the APWPs are consistent and that North China arrived on the northwestern Laurentia margin prior to 1220 Ma. In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwestern Laurentia.

@@

\noindent\textbf{Granite-rhyolite province } A geologic province that comprises widespread Mesoproterozoic rhyolite and granite extending from Labrador, Canada to west Texas, USA.

\noindent\textbf{Grenville orogen } An orogen resulting from the collision Grenvillian orogeny between Laurentia and conjugate continent(s) near the end of the Mesoproterozoic.

@@

\noindent\textbf{juvenile } An adjective referring to rocks formed from melt recently extracted from the mantle.

@@

\noindent\textbf{Nuna } A hypothesized supercontinent interpreted to have formed late in the Paleoproterozoic era and to have broken apart during the Mesoproterozoic.

@@

\noindent\textbf{Rodinia } A hypothesized supercontinent interpreted to have formed late in the Mesoproterozoic era at the time of the Grenvillian orogeny and to have broken apart during the Neoproterozoic.
